sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-07-21 doc/paper-peerj.Rmd
Add peerj version

---
title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
author: "|
  Sean C. Anderson$^{1,2}$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^3$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^4$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada\n|
  $^3$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
  $^4$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
    Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
      keep_tex: true
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\doublespacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \setlength{\parskip}{1.0\baselineskip}
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{lscape}
  \widowpenalty10000
  \clubpenalty10000
abstract: |
    | \noindent 1. Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
    | 2. Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
    | 4. sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
    | Keywords:
      Gaussian Markov random fields (GMRF),
      generalized linear mixed effects models (GLMM),
      INLA,
      SPDE,
      species distribution modelling,
      R package,
      spatio-temporal or spatial-temporal,
      Template Model Builder
---

@@

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```
<!-- sdmTMB: spatial and spatiotemporal random field generalized linear mixed-effects models in R in a fast, flexible, and user-friendly interface -->

# Introduction

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

# Model description

@@

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram), Appendix 1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix 1 for a full description of the model.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

@@

# Introductory example using fish survey data

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

@@

We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Appendix 2).

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->

@@

# Example of a spatially varying coefficient with citizen science data
<!-- # Estimate a spatially varying effect from citizen science data -->

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Appendix 3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NAO unit.

@@

# Model validation and selection

@@

# Package comparisons

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with various limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Appendix 4) and VAST (Appendix 5).

@@

knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
  booktabs = TRUE,
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::footnote(
    general = paste(
      "$^1$Technically possible but non-trivial.",
      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
      "$^3$inlabru but not INLA.",
      "$^4$Spatiotemporal fields as random walk only.",
      "$^5$SPDE approach as in Miller et al. (2019).",
      "$^6$Does have predictive process knots.",
      "$^7$Zero-inflated NB2 only.",
      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
      "$^{10}$Hurdle models possible by fitting components separately."
    ),
    escape = FALSE, threeparttable = TRUE
  )
```

@@

We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
mgcv was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons 95% quantiles across 50 random iterations.
Note the log x and y axes.
<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
VAST is similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
Note that spaMM only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

# Discussion

>>>>

---
title: "sdmTMB: an R package for fast, flexible, and accessible generalized linear mixed effects models with spatial and spatiotemporal random fields"
preprint: false
author:
  - name: Sean C. Anderson
    affiliation: 1
    corresponding: true
    email: sean.anderson@dfo-mpo.gc.ca
  - name: Eric J. Ward
    affiliation: 2
  - name: Philina A. English
    affiliation: 1
  - name: Lewis A. K. Barnett
    affiliation: 3
affiliation:
  - code: 1
    address: Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - code: 2
    address: Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
  - code: 3
    address: Alaska Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
abstract: >
  Geostatistical data---spatially referenced observations of a continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement such that spatial autocorrelation is frequently ignored, or addressed in less ecologically interpretable ways.
  Here, we introduce the R package sdmTMB, which extends the flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB), although penalised likelihood and Bayesian inference (via tmbstan/Stan) are also possible. We describe the statistical model, illustrate the package's use to advance research and monitoring through two case studies, and compare the functionality, speed, and interface to related software.
  The case studies illustrate sdmTMB’s unmatched flexibility and useability by implementing penalised smoothers, non-stationary processes (time-varying and spatially varying relationships), and anisotropy (asymmetry in spatial correlations that commonly occurs along ecological gradients and is not available in INLA/inlabru, the most commonly used packages for this class of models). Modelling of these ecologically ubiquitous, but computationally complex, relationships is further aided by built-in handling of delta/hurdle models, break-point effects, abundance index standardisation, and diagnostic tools including out-of-sample cross validation. Finally, we demonstrate that sdmTMB is 4–16 times faster than INLA or inlabru.
  sdmTMB's accessible interface has already opened this useful class of models to a wider audience within species distribution modelling. We anticipate its application to the full spectrum of novel ecological problems that involve geostatistical data will prove equally fruitful.
output:
  bookdown::pdf_book:
    template: template-peerj.tex
    keep_tex: true
    pandoc_args: "--highlight=haddock"
bibliography: refs.bib
link-citations: yes
csl: peerj.csl
header-includes:
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
---

@@

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

<!-- sdmTMB: spatial and spatiotemporal random field generalized linear mixed-effects models in R in a fast, flexible, and user-friendly interface -->

<!-- File naming format: 'Supplemental [Item] S[number]' e.g. 'Supplemental Data S1'. -->
<!-- Use the following style when citing Supplemental files in the text: -->
<!-- Fig. S1, Table S1, Data S1, Video S1, Article S1, Audio S1. -->

<!-- Introduction -->
<!-- Materials & Methods -->
<!-- Results -->
<!-- Discussion -->
<!-- Conclusions -->
<!-- Acknowledgments -->

# Introduction {-}

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself, for example, by identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

# Model description {-}

@@

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram), Article S1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays at the same rate in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

(ref:diagram-fig) **Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.**
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Article S1 for a full description of the model.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Article S1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in mgcv [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Article S1).
The smoothers in sdmTMB can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution with Stan [@carpenter2017] for Bayesian inference.

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, to impart prior information, to or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

@@

# Introductory example using fish survey data {-}

@@

\noindent
Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
\noindent
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

@@

We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \@ref(fig:pcod-fig)E, Article S2).

(ref:pcod-fig-cap) **Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.**
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (`get_cog()`) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother. These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (`get_index()`). The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
<!-- LB: I wasn't sure which panel the statement about omitting the spatial random field applied to, but moved things around. -->

@@

# Example of a spatially varying coefficient with citizen science data  {-}
<!-- # Estimate a spatially varying effect from citizen science data -->

@@

We found a weak average positive effect of the annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig. \@ref(fig:owl-nao), Article S3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by the NAO.

(ref:owl-nao-fig) **Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.**
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NAO unit.

@@

# Model validation and selection  {-}

@@

# Package comparisons  {-}

There are several R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include spBayes in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with various limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA or inlabru (Article S4) and VAST (Article S5).

@@

knitr::kable(d, format = "latex", caption = "\\textbf{Comparison of functionality between several R packages that can fit geostatistical GLMMs.}",
  booktabs = TRUE,
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::footnote(
    general = paste(
      "$^1$Technically possible but non-trivial.",
      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
      "$^3$inlabru but not INLA.",
      "$^4$Spatiotemporal fields as random walk only.",
      "$^5$SPDE approach as in Miller et al. (2019).",
      "$^6$Does have predictive process knots.",
      "$^7$Zero-inflated NB2 only.",
      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
      "$^{10}$Hurdle models possible by fitting components separately."
    ),
    escape = FALSE, threeparttable = TRUE
  )
```

@@

We ran a simple speed comparison between sdmTMB, inlabru/INLA, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Article S6).
In this test, sdmTMB was the fastest across all mesh resolutions; however, inlabru/INLA and spaMM were less affected by mesh resolution than sdmTMB.
mgcv was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

(ref:timing-fig) **Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.**
Lines represent means and ribbons 95% quantiles across 50 random iterations.
Note the log x and y axes.
<!-- sdmTMB used the default settings and `sdmTMBcontrol(normalize = TRUE)`, which calculates the random effect normalizing constant in R. -->
VAST is similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
Note that spaMM only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

# Discussion {-}

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA will find inlabru approachable.
Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Article S5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find sdmTMB more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages sdmTMB and VAST were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->
